---
title: "Reproducible Research: Course Project 1"
author: "Deon Jacobs"
date: "16 July 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

##Loading and preprocessing the data

```{r}
  library(ggplot2)
  library(dplyr)
  
  #set path where files are located
  setwd("C:/Users/jacobs_d/Documents/R_work/Coursera_work/Reproducible Research/RepData_PeerAssessment1")
  
  #activity data filename
  activityFile <- "activity.csv"
  
  #read csv file with activity data
  activity <-read.csv(activityFile,header = TRUE)
  
  #convert date variable to date format
  activity$date <- as.Date(activity$date,"%Y-%d-%d")
  
  #remove all the NA values from the imported activity dataframe
  no_NAs <- activity[complete.cases(activity),]
```

##What is mean total number of steps taken per day?

###Calculate the total number of steps taken per day:

```{r}
  
  #calculate the step total count per day
  sumStepsPerDay <- group_by(no_NAs,date) %>% summarise(sum_steps=sum(steps))
  
  #generate histogram plot of the sum of the daily steps
  g <- ggplot(sumStepsPerDay, aes(x=sum_steps))
  
  g + geom_histogram(binwidth=2500,colour="black") + 
  scale_x_continuous(breaks = c(seq(0,35000,by=2500))) + theme_bw(base_family="Times")
```

###Mean and median of the total number of steps taken per day:  
```{r}
  #calculate the mean and median step count per day for all intervals
  statsStepsPerDay <- round(summarise(sumStepsPerDay,Mean = mean(sum_steps), Median = median(sum_steps)),0)
  #display daily mean values
  knitr::kable(statsStepsPerDay)
```

##What is the average daily activity pattern?  

###Five minute interval vs. averaged daily step count across all days  
```{r}
  #Calculate average steps over the intervals for all days
  fiveMinIntDailyAvg <- group_by(no_NAs,interval) %>% summarise(avg_steps=mean(steps))
  
  #Plot line time series view 
  g <- ggplot(fiveMinIntDailyAvg,aes(x=interval,y=avg_steps))
  g + geom_line(size=1.5) + scale_x_continuous(breaks = c(seq(0,2400,by=250))) + theme_bw(base_family="Times") +
    labs(x="Interval",y="Average Steps All Days")
```

### Five minute interval which on average across all the days in the dataset contains the maximum number of steps

```{r}
  maxFiveMinInt <- fiveMinIntDailyAvg[fiveMinIntDailyAvg$avg_steps == max(fiveMinIntDailyAvg$avg_steps),]
  knitr::kable(maxFiveMinInt)
```

##Imputing missing values  

###Number of NA rows in the activity dataset:

```{r}
numOfNAs <- sum(is.na(activity$steps))
dfNumNAs <- data.frame(numOfNAs)
knitr::kable(dfNumNAs)
```
###Strategy for Imputing Missing Values  

1. Calculate mean across all intervals
2. Substitute NA step values with calculated interval means

```{r}
activityNew <- activity
  NAs <- is.na(activityNew$steps)
  avgInterval <- tapply(activityNew$steps, activityNew$interval, mean, na.rm=TRUE, simplify=TRUE)
  activityNew$steps[NAs] <- avgInterval[as.character(activityNew$interval[NAs])]
```
3. Check if any NAs still exist in new activity dataset:

```{r}
numOfNAs <- sum(is.na(activityNew$steps))
dfNumNewNA <- data.frame(numOfNAs)
knitr::kable(dfNumNewNA)
```
###Calculate new mean average steps per day Histogram, Median and Mean values with New activity dataset
```{r}
  
  #calculate the step total count per day
  sumStepsPerDay <- group_by(activityNew,date) %>% summarise(sum_steps=sum(steps))
  
  #generate histogram plot of the sum of the daily steps
  g <- ggplot(sumStepsPerDay, aes(x=sum_steps))
  
  g + geom_histogram(binwidth=2500,colour="black") + 
  scale_x_continuous(breaks = c(seq(0,35000,by=2500))) + theme_bw(base_family="Times")
```
###Mean and median of the total number of steps taken per day:
```{r}
  #calculate the mean and median step count per day for all intervals
  statsStepsPerDay <- round(summarise(sumStepsPerDay,Mean = mean(sum_steps), Median = median(sum_steps)),0)
  #display daily mean values
  knitr::kable(statsStepsPerDay)
```

###Effect of imputing missing values  

1. Imputing missing values do alter the statistical results calculated from the activity data set. 
2. By imputing missing values the mean value of the daily sum of steps increases, implying an increase in daily steps.

##Are there differences in activity patterns between weekdays and weekends?

###Create weeday clasiffication variable from new activity dataset

```{r}
activityDate <- mutate(activityNew, dayClass = ifelse(weekdays(date) == "Saturday" | weekdays(date) == "Sunday", "weekend", "weekday"))
activityDate$dayClass <- as.factor(activityDate$dayClass)
```
###Panel plot of weekday vs. weekend 
```{r}
  #Calculate average steps over the intervals for all days
  fiveMinIntDailyAvg <- group_by(activityDate,interval,dayClass) %>% summarise(avg_steps=mean(steps))
  
  #Plot line time series view 
  g <- ggplot(fiveMinIntDailyAvg,aes(x=interval,y=avg_steps))
  g + geom_line(size=1.5) + scale_x_continuous(breaks = c(seq(0,2400,by=250))) + theme_bw(base_family="Times") +
    labs(x="Interval",y="Average Steps All Days") + facet_grid(dayClass~.)
```  

1. A clear difference in weekday vs. weekend step count characteristic is observerd from the above panel plot.
